Adaptive Disk Storage for Interaction Graphs

### 2014-08-07 docs/cidr2015/text/main.tex
smoother introduction

\numberofauthors{2} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Robert Soul\'{e}\\
       \affaddr{University of Lugano}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{robert.soule@usi.ch}
% 2nd. author
\alignauthor
Bu\u{g}ra Gedik\\
       \affaddr{Bilkent University}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{bgedik@cs.bilkent.edu.tr}
}

@@

Query workload and disk layout have a significant impact on database
performance.  For table-based, relational databases, this fact has led database
designers to develop alternative approaches for storage layout: row-oriented
storage is more efficient when queries access many attributes from a small
number of records, and column-oriented storage is more efficient when queries
access a small number of attributes from many records.

Recently, there has been increased interest in graph databases in general, and
\emph{interaction graph} databases in particular.  An \emph{interaction graph}
is an append-only graph, where new edges and vertices are added as time
progresses.  Thus, interaction graph encode the relationships over time between
entities in their structure, making them ideal for use in social networking,
telecommunications, or modeling the world-wide web.

Unfortunately, although interaction graph databases, like relational databases,
are the target of diverse query workloads, there is no clear correspondence to a
row-oriented or column-oriented storage layout.
Typical interaction graphs have data (i.e., \emph{attributes} ) associated with
their vertices and edges. Most algorithms access this data as they traverse the
graph.

In an interaction graph database organized as blocks of temporal neighbor lists
(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
vertex attributes locally. For  instance, if the edge properties were kept separately
in a relational table, there will be almost no benefit to the locality
optimizations performed for block organization (discussed in Gedik et
al.~\cite{gedik14}, as we would have to go back and forth between the disk
blocks to access the edge attributes).%

On the other hand, putting all the edge attributes into the disk blocks
containing the graph structure is expected to cause significant overhead when
only a few of these attributes are read. This is somewhat similar to the

% However, not all of the data is
% accessed by all the queries. Typically, there are correlations among the
% attributes accessed by different queries, such as Q1 and Q5 accessing
% attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%

Typically, a graph database stores entities as vertices,
relationships as edges, and additional information in labels that may be
associated with either vertices or edges.

They store entities as vertices,
relationships as edges, and additional information in labels that may be
associated with either vertices or edges.

An \emph{interaction graph} is an append-only graph, where new edges and
vertices are added as time progresses.%

Typical interaction graphs have data associated with their vertices and edges.
We call this data \emph{attributes} or \emph{properties}. Most algorithms
access this data as they traverse the graph. However, not all of the data is
accessed by all the queries. Typically, there are correlations among the
attributes accessed by different queries, such as Q1 and Q5 accessing
attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%

% In an interaction graph database organized as blocks of temporal neighbor lists
% (as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
% vertex properties locally. For  instance, if the edge properties are kept away
% in a relational table, there will be almost no benefit to the locality
% optimizations performed for block organization (discussed in Gedik et
% al.~\cite{gedik14}, as we would have to go back and forth between the disk
% blocks to access the edge attributes).%

On the other hand, putting all the edge attributes into the disk blocks
containing the graph structure is expected to cause significant overhead when
only a few of these attributes are read. This is somewhat similar to the 
problem with row-oriented databases, where the entire row needs to be accessed
from the disk despite the query needing only a small fraction of it.
Unfortunately, there is no clear correspondence to a column-oriented database
layout for the graph databases.%

An important characteristic of interaction graphs are that they are temporal.
As such, the co-access correlations for the attributes can be different for
different temporal regions. It might also be unknown at the insertion time and
may be discovered later depending on the workload. This points to the need for
adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
and HYRISE~\cite{grund10}).%

I imagine a solution to this problem, which I name the `rail layout'. The idea
is to start with large blocks that contain the entire data. As we learn about
the access properties for different time regions, we might start splitting
such blocks into smaller blocks that run parallel to each other, almost like
having two or more graph databases for certain time regions, each containing a
 different subset of the attributes, but with a link between them in case a
query needs to access both.

@@

\subsubsection{Heuristic Solution}

\paragraph*{Approach.$\,$}
TODO

\begin{algorithm}[ht]
\scriptsize
\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
\label{alg:overlappingP}
\KwData{$B$: block, $Q$: set of queries}
$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
      $j\leftarrow -1$ \tcp*{Best partition assignment}
      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
            $j\leftarrow i$\tcp*{Update the best partition}
         }
         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
      }
      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
   }
   \tcp{If solution is feasable and has lower cost}
   \If{$H(R, B, Q)\geq\alpha \wedge L(R, B, Q)<c^*$}{
      $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
      $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
   }
}
\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
\end{algorithm}

@@

\subsubsection{Heuristic Solution}

\paragraph*{Approach.$\,$} We start the algorithm with a 
 partitioning based on what queries we have seen.
Every query gets its own sub-block. This is the``ideal'' partitioning,
because the I/O cost would be minimized for every query that we would
have seen. As input to the algorithm, we provide a function $\dist$, which
is used to compute the distance between two blocks.

$$
\dist: block \rightarrow block \rightarrow int
$$

Note that the $\dist$ function can be chosen arbitrarily.  Examples of useful
$\dist$ functions are the frequencies with which the queries are accessed, or the Jaccard index, which measures the set-similarity of
the attributes.

Using this function, the algorithm iteratively combines the two partitions that
are closest together. Algorithm~\ref{alg:closestBF} shows the pseudocode for a brute-force approach
  to finding the closest paris that runs in $O(n^2)$ time. However, there is
  also a divide-and-conquer algorithm~\cite{cormen01} for solving closest pair
  of points problem in $O(n \log n)$.
After each combination of partitions, the algorithm calculate the storage cost
for the partitioning. The algorithm stops when the  storage cost is below some
specified threshold.  The result is the block partitioning.

\begin{algorithm}[ht]
\SetKwFunction{closestPair}{closestPair}
\scriptsize
\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
\label{alg:closestPair}
\KwData{$B$: block, $Q$: set of queries}
$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}

\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
 $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
}

\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
  \text{max}$}{
$\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
 $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - \{x, y\}$ \;
  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  \cup \{ x \cup y \}$ \;
}
\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
\end{algorithm}

 \begin{algorithm}[ht]
\SetKwFunction{length}{length}
\SetKwFunction{dist}{dist}
 \scriptsize
 \caption{Brute-force algorithm for closest points in $O(n^2)$.}
 \label{alg:closestBF}
$minDist  \gets \infty $ \;
\For{$i \gets 1$ \textbf{to} $\length(P)-1$} {
\For{$j \gets i+ 1$ \textbf{to} $\length(P)$} {

  $p \gets P[i]$ \;
  $q \gets P[j]$ \;
  \If{$\dist(p,q) < minDist$} {
    $minDist \gets \dist(p,q)$\;
    $closestPair \gets (p,q)$\;
  }
}
}
\Return $closestPair$
% \KwData{$B$: block, $Q$: set of queries}
% $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
% \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
%  $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
% }
% \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
%   \text{max}$}{
% $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
%  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
%   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
% \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
 \end{algorithm}

\section{Related Work}
\section{Conclusion}

>>>>

\numberofauthors{2}

\author{
% 1st. author
\alignauthor
Robert Soul\'{e}\\
       \affaddr{University of Lugano}\\
       \affaddr{Faculty of Informatics}\\
       \affaddr{Lugano 6904, Switzerland}\\
       \email{robert.soule@usi.ch}
% 2nd. author
\alignauthor
Bu\u{g}ra Gedik\\
       \affaddr{Bilkent University}\\
       \affaddr{Department of Computer Engineering}\\
       \affaddr{Bilkent, Ankara 06800, Turkey}
       \email{bgedik@cs.bilkent.edu.tr}
}

@@

Graph databases, and \emph{interaction graph} databases in
particular, are a critical component for many of today's most popular
applications.  Graph databases store entities as vertices, relationships as
edges, and have data associated with either vertices or edges, called
\emph{attributes} or \emph{properties}.  Interaction graphs differ from general
graphs in that they are \emph{append-only}, and edges are assigned a
\emph{temporal value}.  Thus, interaction graphs encode the relationships over
time between entities in their structure, making them ideal for use in social
networking, telecommunications, or modeling the world-wide web.

To query an interaction graph, most algorithms traverse the graph structure to
access the relevant attributes.  Frequently, there are correlations among the
attributes accessed by different queries. For example, queries $q_1$ and $q_5$
might access attributes $a_1$ and $a_2$, while queries $q_2$, $q_3$ and $q_4$
access attributes $a_3$ and $a_4$. Because interaction graphs are temporal, the
co-access correlations for the attributes can vary for different
temporal regions.  Moreover, the co-access correlations might be unknown at the
insertion time, but be discovered later, when the workload is known.

It is widely recognized that query workload and disk layout have a significant
impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
table-based, relational databases, this fact has led database designers to
develop alternative approaches for storage layout: row-oriented storage is more
efficient when queries access many attributes from a small number of records,
and column-oriented storage is more efficient when queries access a small number
of attributes from many records~\cite{stonebraker05}.  Unfortunately, although
interaction graph databases, like relational databases, are the target of
diverse query workloads, there is no clear correspondence to a row-oriented or
column-oriented storage layout.

Typically, an interaction graph database is organized as blocks of temporal
neighbor lists.  Attributes can be stored in two ways, either separately (e.g., in a relational
table), or locally with the temporal neighbor lists.  If they are stored
separately, then the graph database cannot take advantage of locality
optimizations performed for block organization.  As discussed in Gedik et
al.~\cite{gedik14}, the database must go back and forth between the disk blocks
to access the edge attributes.  On the other hand, if attributes are stored
locally in the disk blocks containing the graph structure, then there can be
significant overhead due to disk I/O if only a few attributes are needed to
answer a query.

This paper presents an algorithm for adaptively optimizing disk block storage
for interaction graphs. The key idea behind the approach is a novel storage
scheme, called a \emph{rail layout}.  Initially, attributes are stored with
their neighbor lists in large blocks. Over time, the blocks are split into
smaller-blocks that contain the neighbor lists, but only subsets of the
attributes, depending on the query workload. Intuitively, this is almost like
having two or more graph databases for certain time regions, each containing a
different subset of the attributes, but with a link between them in case a query
needs to access both.

% This points to the need for
% adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
% and HYRISE~\cite{grund10}).%

@@

@@

\section{Evaluation}
\section{Related Work}

\begin{alltt}\scriptsize
- H2O \cite{alagiannis14}
- HYRISE~\cite{grund10}).%
\end{alltt}
\section{Conclusion}
